Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-11-15 manuscript/manuscript.Rmd
Cleaning up discussion, adding refs, final edits

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 5). We found that a majority of the metabolites (Clusters 1 and 2, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 5). Cluster 2 had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (Pearson's $r=0.851$) that was not present in the cyclone ($r=0.013$).

@@

Almost all large phytoplankton classes in the 2017 eddy pair were enriched in the cyclone DCM relative to the anticyclone (particularly prymnesiophytes, dinoflagellates, diatoms, and green algae), while *Prochlorococcus* was significantly less abundant and *Synechococcus* was approximately equal [@Barone2022]. Thus, the metabolite data presented here does not help distinguish between large phytoplankton classes but instead provides additional cues about the shift from a metabolome dominated by cyanobacteria producers into one dominated by eukaryotic organisms. This shift occurs naturally over the life cycle of an eddy, with distinctions typically drawn between the eddy intensification, its mature phase, and eventual decay [@Rii2008; @McGillicuddy2016; @Kuhlisch2023; @Barone2022]. Both cruises occurred in weakening cyclones and mature anticyclones, limiting our ability to make direct conclusions about all eddy life cycle phases, but the general alignment of the results presented here with our initial hypotheses strengthens our confidence in metabolites as links between eddy biogeochemistry and their biological shifts anywhere.

>>>>

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 5). We found that a majority of the metabolites (Clusters 1 and 2, 63% of the total) fell into clusters with larger peak areas in the cyclone while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 5). Cluster 2 had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (Pearson's $r=0.851$) that was not present in the cyclone ($r=0.013$).

@@

Almost all large phytoplankton classes in the 2017 eddy pair were enriched in the cyclone DCM relative to the anticyclone (particularly prymnesiophytes, dinoflagellates, diatoms, and green algae), while *Prochlorococcus* was significantly less abundant and *Synechococcus* was approximately equal [@Barone2022]. Thus, the metabolite data presented here does not help distinguish between large phytoplankton classes but instead provides additional cues about the shift from a metabolome dominated by cyanobacteria producers into one dominated by eukaryotic organisms. This shift occurs naturally over the life cycle of an eddy, with distinctions typically drawn between the eddy intensification, its mature phase, and eventual decay [@Rii2008; @McGillicuddy2016; @Kuhlisch2023; @Barone2022]. The initial introduction of inorganic nutrients during intensification of cyclonic eddies followed by diapycnal mixing can still stimulate the DCM community without large effects on integrate primary productivity or export even months after formation, as seen here. Both cruises occurred in weakening cyclones and mature anticyclones, limiting our ability to make direct conclusions about all eddy life cycle phases. However, the general alignment of the results presented here with our initial hypotheses strengthens our confidence in metabolites as links between eddy biogeochemistry and the associated biological shifts anywhere.
